---
title: "P3 Kids - experiment look at whether we can use SRS cross-sectional approach"
subtitle: "`r knitr::current_input(dir = TRUE)`"
author: "Mark Jones"
date: "`r Sys.time()`"
editor_options:
  chunk_output_type: console
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    fig_caption: yes
---


# Preamble

Test whether you can turn a large dataset comprising clusters and subjects with repeat measures on the subjects into a cross-sectional dataset by simple random sampling and still get reasonably good recovery of the parameters of interest.

```{r, echo = F}
suppressPackageStartupMessages(library(compiler))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(doParallel))
suppressPackageStartupMessages(library(foreach))

ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(text = element_text(size = 10))
ggplot2::theme_update(legend.position = "top")
# ggplot2::theme_update(legend.title = element_blank())
ggplot2::theme_update(axis.text.x = element_text(size = 10))
ggplot2::theme_update(axis.text.y = element_text(size = 10))


var_e <- (pi^2)/3

inv_logit <- cmpfun(function(x){
  return(exp(x)/(1+exp(x)))
})
odd <- cmpfun(function(p){
  return(p/(1-p))
})
prob <- cmpfun(function(odd){
  return(odd/(1+odd))
})

```

# Data generation assumptions

Eight clusters, each with 1000 subjects each of which have 5 observations.

```{r}
# parameters
getpars <- function(){
  pars <- list()
  pars$p0 <- 0.45
  pars$sig_hosp <- sqrt(0.1)
  pars$sig_subj <- sqrt(0.6)
  pars$t <- 0:4
  pars$n_hosp <- 8
  pars$n_subj <- 5000
  pars$tx <- 0.7
  pars$nsim <- 1000
  pars$cluster_size <- 400
  pars
}

```



```{r, echo = T}
gendat <- cmpfun(function(mypars){    # 1.77
  
  # this is a ridiculously big dataset
  # ids for patients in hospitals: sequential from 1 to hosp N
  d0 <- data.table::CJ(1:mypars$n_hosp,
                       1:mypars$n_subj,
                       mypars$t)
  names(d0) <- c("hosp_id", "subj_id", "time")
  
  d0$id <- interaction(d0$hosp_id, d0$subj_id)

  # baseline prob (grand mean)
  d0$p0 <- mypars$p0
  
  ds <- data.frame(hosp_id = 1:mypars$n_hosp, 
                   tx_start = rep(1:(mypars$n_hosp/2), each = 2))
  d0 <- merge(d0, ds, by = "hosp_id")
  d0$tx_active <- as.numeric(d0$time >= d0$tx_start)
  
  hosp_int <- rnorm(mypars$n_hosp, 0, mypars$sig_hosp)
  l <- lapply(1:length(hosp_int), 
              function(x){ 
                data.table(subj_int = 
                             rep(rnorm(mypars$n_subj, 0, mypars$sig_subj),
                                 each = length(mypars$t)),
                           hosp_int = hosp_int[x])
                })

  d0 <- cbind(d0, rbindlist(l))
  
  d0$eta <- log(odd(d0$p0)) + mypars$tx * d0$tx_active + 
    d0$hosp_int + d0$subj_int
  d0$p <- inv_logit(d0$eta)
  d0$y <- rbinom(nrow(d0), 1, prob = d0$p)

  d0
})

```

Generate data with selected parameter values. 

```{r, echo = T}
mypars <- getpars()
d <- gendat(mypars)
l1 <- glmer(y ~ tx_active + (1|hosp_id/subj_id) , 
            data = d, family = binomial)
summary(l1)
sjstats::icc(l1, adjust = F)
```


```{r, echo = F}
icc1 <- sjstats::icc(l1, adjust = F)
dpar <- data.table(term = c("intercept", "trt", 
                            "var_subj", "var_hosp"))
dpar$true_val <- c(log(odd(mypars$p0)), mypars$tx, 
                   mypars$sig_subj^2, mypars$sig_hosp^2)
dpar$true_val <- round(dpar$true_val, 3)
v <-  as.data.frame(VarCorr(l1))
dpar$fit_val <- c(fixef(l1), v$vcov)
dpar$fit_val <- round(dpar$fit_val, 3)

dpar %>%
  kable( col.names = c("Term", "True Value", "Fitted")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

Plot this single simulated dataset.

```{r, echo = F, fig.height=6, fig.width=6}
# produce proportion with event
dfig <- d[, .(prop = mean(y)),
          keyby = .(hosp_id, time, tx_active, tx_start)]

# We have five timepoints measured. Just extrapolate the last one
# by hosp and clinic
dfig2 <- dfig[, .SD[.N], by=.(hosp_id)]
dfig2$time <- 5
dfig <- rbind(dfig, dfig2)

# for the start lines
ds1 <- data.table(hosp_id = 1:mypars$n_hosp,
                  tx_start = rep(1:(mypars$n_hosp/2), each = 2))
                        
ggplot(dfig, aes(x = time, y = prop, 
                 group = paste0(hosp_id),
                 colour = paste0(hosp_id)))+
  geom_point(size = 0.4, alpha = 0.2) +
  geom_step() +
  scale_color_discrete("Hospital")+
  scale_x_continuous("Years from start of study", lim = c(0, 5))+
  scale_y_continuous("Proportion vaccinated", lim = c(0, 1), breaks = c(0, 0.5, 1)) +
  facet_wrap(paste0("Start tx ", tx_start) ~ .) +
  geom_vline(data = ds1, aes(xintercept = tx_start), linetype = 2)

```

Simulation using full dataset.

```{r, eval = F}
starttime <- Sys.time()
cl <- makeCluster(parallel::detectCores() - 2, outfile="")
registerDoParallel(cl)
# registerDoSEQ()

mypars <- getpars()

results <- foreach(i = 1:mypars$nsim,
                   .errorhandling = 'pass',
                   #.export = c("inv_logit", "prob", "odd"),
                   .packages=c("lme4", "data.table", "compiler", "sjstats")
                   #.options.snow=opts,
                   ) %dopar%{
                     
  if(i %% 100 == 0) message("i = ", i)
  d <- gendat(mypars)
  l1 <- glmer(y ~ tx_active + (1|hosp_id/subj_id) , 
              data = d, family = binomial)

  if(sjstats::is_singular(l1)){
    return(data.table(idsim = i, Var1 =0,Var2 =0,value=0))
  }
  
  m1 <- data.table::melt(summary(l1)$coef[,1:2])
  m2 <- data.table::melt(as.data.frame(VarCorr(l1))[,c(1,4)], 
                         id.vars = "grp" )
  cbind(id = i, rbindlist(list(m1, m2)))
}

stopCluster(cl)
endtime <- Sys.time()
difftime(endtime, starttime, units = "hours")
rdsfilename <- paste0("res-expt1-", 
                      format(Sys.time(), "%Y-%m-%d-%H-%M-%S"), ".RDS")
saveRDS(list(results = results,
             mypars = mypars, 
             starttime = starttime,
             endtime = endtime ), rdsfilename)
```

Simulation using dataset constructed with SRS.

```{r, eval = F}
starttime <- Sys.time()
cl <- makeCluster(parallel::detectCores() - 2, outfile="")
registerDoParallel(cl)
# registerDoSEQ()
mypars <- getpars()
results <- foreach(i = 1:mypars$nsim,
                   .errorhandling = 'pass',
                   #.export = c("inv_logit", "prob", "odd"),
                   .packages=c("lme4", "data.table", "compiler", "sjstats")
                   #.options.snow=opts,
                   ) %dopar%{
 
  if(i %% 100 == 0) message("i = ", i)
  d <- gendat(mypars)
  dsub <- data.table()
  for(k in 1:mypars$n_hosp){
    subjs <- d[hosp_id == k & time == 0, subj_id]
    idx <- sample(1:length(subjs), 
                  size = mypars$cluster_size, 
                  replace = T)

    for(z in mypars$t){
      dsub <- rbind(dsub, d[hosp_id == k & time == z, ][idx,])
    }
    
  }

  
  # assumes that the samples are cross-sectional from srs
  l1 <- glmer(y ~ tx_active + (1|hosp_id/subj_id) , 
              data = dsub, family = binomial)

  if(sjstats::is_singular(l1)){
    return(data.table(idsim = i, Var1 =0,Var2 =0,value=0))
  }
  
  m1 <- data.table::melt(summary(l1)$coef[,1:2])
  m2 <- data.table::melt(as.data.frame(VarCorr(l1))[,c(1,4)], 
                         id.vars = "grp" )
  cbind(id = i, rbindlist(list(m1, m2)))
}

stopCluster(cl)
endtime <- Sys.time()
difftime(endtime, starttime, units = "hours")
rdsfilename <- paste0("res-expt2-", 
                      format(Sys.time(), "%Y-%m-%d-%H-%M-%S"), ".RDS")
saveRDS(list(results = results,
             mypars = mypars, 
             starttime = starttime,
             endtime = endtime ), rdsfilename)
```



```{r, eval = F, echo = F}
fnames <- c("out/res-expt1-2019-05-02-04-01-09.RDS",
            "out/res-expt2-2019-05-02-10-40-17.RDS")

lres <- vector("list", length(fnames))
for(z in 1:length(fnames)){
  res <- readRDS(fnames[z])
  dfig <- rbindlist(lapply(1:1000, function(x) res$results[[x]]))
  lres[[z]]$n_singular <- sum(dfig$value == 0)
  dtab <- dfig[value != 0][, mean(value), by = .(Var1, Var2)]
  
  lres[[z]]$intercept <- sprintf("%.3f (%.3f)", 
                                 dtab[Var1 == "(Intercept)" & 
                                        Var2 == "Estimate", V1],
                                 dtab[Var1 == "(Intercept)" & 
                                        Var2 == "Std. Error", V1])
  
  lres[[z]]$tx_active <- sprintf("%.3f (%.3f)", 
                                 dtab[Var1 == "tx_active" & 
                                        Var2 == "Estimate", V1],
                                 dtab[Var1 == "tx_active" & 
                                        Var2 == "Std. Error", V1])
  
  lres[[z]]$tx_active <- sprintf("%.3f (%.3f)", 
                                 dtab[Var1 == "tx_active" & 
                                        Var2 == "Estimate", V1],
                                 dtab[Var1 == "tx_active" & 
                                        Var2 == "Std. Error", V1])
  
  lres[[z]]$subj_id <- sprintf("%.3f", 
                                 dtab[Var1 == "subj_id:hosp_id" & 
                                        Var2 == "vcov", V1])
  
  lres[[z]]$hosp_id <- sprintf("%.3f", 
                                 dtab[Var1 == "hosp_id" & 
                                        Var2 == "vcov", V1])
  
}

dtab <- rbindlist(lres)
```

Now compare the results. The following table suggests that the approach is probably reasonable.

```{r, eval = F}
dpar <- data.table(term = c("intercept", "trt", 
                            "var_subj", "var_hosp"))
dpar$true_val <- c(log(odd(mypars$p0)), mypars$tx, 
                   mypars$sig_subj^2, mypars$sig_hosp^2)
dpar$true_val <- round(dpar$true_val, 3)

dpar$fit_val_dat_all <- as.matrix(dtab[1, 2:5])[1,]
dpar$fit_val_dat_sub <- as.matrix(dtab[2, 2:5])[1,]

dpar %>%
  kable( col.names = c("Term", "True Value", 
                       "Fitted (all)", "Fitted (subset)")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```


References

For nested effects.
https://www.muscardinus.be/2017/07/lme4-random-effects/


