---
title: "P3 Kids - power and sample size simulation"
subtitle: "`r knitr::current_input(dir = TRUE)`"
author: "Mark Jones"
date: "`r Sys.time()`"
editor_options:
  chunk_output_type: console
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    fig_caption: yes
---


# Preamble

Test whether you can turn a large dataset comprising clusters and subjects with repeat measures on the subjects into a cross-sectional dataset by simple random sampling.


```{r, echo = F}
suppressPackageStartupMessages(library(compiler))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(doParallel))
suppressPackageStartupMessages(library(foreach))

ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(text = element_text(size = 10))
ggplot2::theme_update(legend.position = "top")
# ggplot2::theme_update(legend.title = element_blank())
ggplot2::theme_update(axis.text.x = element_text(size = 10))
ggplot2::theme_update(axis.text.y = element_text(size = 10))


var_e <- (pi^2)/3

inv_logit <- cmpfun(function(x){
  return(exp(x)/(1+exp(x)))
})
odd <- cmpfun(function(p){
  return(p/(1-p))
})
prob <- cmpfun(function(odd){
  return(odd/(1+odd))
})

```


# Data generation assumptions

Eight clusters, each with 1000 subjects each of which have 5 observations.


```{r, echo = T}
gendat <- cmpfun(function(p0 = 0.45, 
                          sig_hosp = sqrt(0.1),
                          sig_subj = sqrt(1.77)){
  # Timepoints
  t <- 0:4
  n_hosp <- 8
  n_subj <- 1000
  
  # this is a ridiculously big dataset
  # ids for patients in hospitals: sequential from 1 to hosp N
  d0 <- data.table::CJ(1:n_hosp,
                       1:n_subj,
                       t)
  names(d0) <- c("hosp_id", "subj_id", "time")

  # baseline prob (grand mean)
  d0$p0 <- p0
  
  ds <- data.frame(hosp_id = 1:8, tx_start = rep(1:4, each = 2))
  d0 <- merge(d0, ds, by = "hosp_id")
  d0$tx_active <- as.numeric(d0$time >= d0$tx_start)
  
  hosp_int <- rnorm(n_hosp, 0, sig_hosp)
  l <- lapply(1:length(hosp_int), 
              function(x){ 
                data.table(subj_int = rep(rnorm(n_subj, hosp_int[x], sig_subj), each = length(t)),
                           hosp_int = hosp_int[x])
                })

  d0 <- cbind(d0, rbindlist(l))
  
  d0$eta <- log(odd(d0$p0)) + 0.7 * d0$tx_active + d0$hosp_int + d0$subj_int
  d0$p <- inv_logit(d0$eta)
  d0$y <- rbinom(nrow(d0), 1, prob = d0$p)

  d0
})

```

Generate data with selected parameter values. 

```{r, echo = F}
d <- gendat()
l1 <- glmer(y ~ tx_active + (1|hosp_id/subj_id) , 
            data = d, family = binomial)
summary(l1)$coef
```

Simulation using full dataset.

```{r}
nsim <- 1000
starttime <- Sys.time()
cl <- makeCluster(parallel::detectCores() - 2, outfile="")
registerDoParallel(cl)
# registerDoSEQ()

results <- foreach(i = 1:nsim,
                   .errorhandling = 'pass',
                   #.export = c("inv_logit", "prob", "odd"),
                   .packages=c("lme4", "data.table", "compiler", "sjstats")
                   #.options.snow=opts,
                   ) %dopar%{
 
  d <- gendat()
  l1 <- glmer(y ~ tx_active + (1|hosp_id/subj_id) , 
              data = d, family = binomial)

  if(sjstats::is_singular(l1)){
    return(data.table(idsim = i, Var1 =0,Var2 =0,value=0))
  }
  
  m1 <- data.table::melt(summary(l1)$coef[,1:2])
  m2 <- data.table::melt(as.data.frame(VarCorr(l1))[,c(1,4)], id.vars = "grp" )
  cbind(id = i, rbindlist(list(m1, m2)))
}

stopCluster(cl)
endtime <- Sys.time()
difftime(endtime, starttime, units = "hours")
saveRDS(results, "res_expt1.RDS")
```

Simulation using dataset constructed with SRS.

```{r}
starttime <- Sys.time()
cl <- makeCluster(parallel::detectCores() - 2, outfile="")
registerDoParallel(cl)
# registerDoSEQ()

results <- foreach(i = 1:nsim,
                   .errorhandling = 'pass',
                   #.export = c("inv_logit", "prob", "odd"),
                   .packages=c("lme4", "data.table", "compiler", "sjstats")
                   #.options.snow=opts,
                   ) %dopar%{
 
  d <- gendat()
  dsub <- data.table()
  for(j in 0:4){
    idxmax <- d[time == j, .N]
    idx <- sample(1:idxmax, size = 150, replace = T)
    dsub <- rbind(dsub, d[time == j, ][idx,])
  }
  
  # assumes that the samples are cross-sectional from srs
  l1 <- glmer(y ~ tx_active + (1|hosp_id) , 
              data = dsub, family = binomial)

  if(sjstats::is_singular(l1)){
    return(data.table(idsim = i, Var1 =0,Var2 =0,value=0))
  }
  
  m1 <- data.table::melt(summary(l1)$coef[,1:2])
  m2 <- data.table::melt(as.data.frame(VarCorr(l1))[,c(1,4)], id.vars = "grp" )
  cbind(id = i, rbindlist(list(m1, m2)))
}

stopCluster(cl)
endtime <- Sys.time()
difftime(endtime, starttime, units = "hours")
saveRDS(results, "res_expt2.RDS")
```

Now compare the results.